---
title: "Good Programming Practices Exercises"
author: "N. Schenk"
date: "2022-11-09"
output: html_document
---
# Good Programming Practices

477757-HS2022-0-R Coding Club, Application for Ecology. Examples and exercises adapted from [Synthesis helpdesk Good Programming Practices](https://github.com/biodiversity-exploratories-synthesis/Synthesis_courses_workshops/tree/main/Help_desk_Fall_2021). Written by Noelle Schenk, 09.11.22.

This file contains examples of practices to imporove the readability of our code. Feel free to use this file to play around with the different examples yourself.


## Dependencies
```{r, eval = F}
library(data.table)
library(reshape2)
library(tidyr)
library(styler)

path_to_data <- "example_data_synthesis_helpdesk/"
path_to_output <- "R_outputs_and_plots/"

set.seed(12) # always generate the same random numbers, every time you run this script
```

## Good Coding practices

### Keep it simple!

Use the lines and indenting.
```{r, eval = F}
# bad example : all in one line
info_data <- data.table::fread(paste(path_to_data, "/info_data.csv", sep = ""), header=T)
#
# good example : use many lines
info_data <- data.table::fread(
  paste(path_to_data, 
        "/info_data.csv", 
        sep = ""), 
  header=T)
```

Add structure to your file
```{r, eval = F}
# add structure to your file ####
# if you add four "#" or "-" in the end of the title, RStudio will allow
#   you to fold in and out the section
# second example ----

# you can even generate nice titles which are easy to find when scrolling around :

# # # # # # # # # # #
#                   #
#      TITLE        # ----
#                   #
# # # # # # # # # # #
```


### Use a code style

Examples taken from the [tidyverse code style](http://adv-r.had.co.nz/Style.html and https://style.tidyverse.org/syntax.html#long-lines)
```{r, eval = F}
# Good
average <- mean(feet / 12 + inches, na.rm = TRUE)
# Bad
average<-mean(feet/12+inches,na.rm=TRUE)

# Good
do_something_very_complicated(
  something = "that",
  requires = many,
  arguments = "some of which may be long"
)
# Bad
do_something_very_complicated("that", requires, many, arguments,
                              "some of which may be long"
)

# Example use of styler with RStudio
# 1. copy bad example from above
# 2. Select the piece of code you wish to style
# 3. Go to RStudio Menu bar above > Addins > search for "styler" > click "Style selection"
average<-mean(feet/12+inches,na.rm=TRUE)
OneFile <- c(1,2,3) # styler does not correct variable names in CamelCase
```


## Comment your code

Better too much commenting than too little.
```{r, eval = F}
# Bad example : comment the obvious
# read table example.csv
info_data <- data.table::fread("example.csv", header = T)

# Good example : comment the unexpected
# use package data.table to read in very large table "example.csv"
info_data <- data.table::fread("example.csv", header = T)
```


### Make a script header

A header contains (at least) :
- purpose of the script
- date and creator name
- requirements/ dependencies

Either in Markdown (see the script header of this file) or in an R script :

```{r, eval = F}
# # # # # # # # # # # # # # # # # ##
#                                  #
# GOOD PROGRAMMING PRACTICES       #
#                                  #
# # # # # # # # # # # # # # # # # ##
# Biodiversity Exploratories Synthesis R helpdesk
# Noëlle Schenk 05.02.2021
#
# this file contains examples of practices to improve the
# readability of your code.
#
# Dependencies : 
library(data.table)
library(reshape2)
library(tidyr)
library(styler)

path_to_data <- "example_data_synthesis_helpdesk/"
path_to_output <- "R_outputs_and_plots/"

set.seed(12) # always generate the same random numbers, every time you run this script
```



### give meaningful names

Good names reduce comments
```{r, eval = F}
time_tolerance <- 60 # seconds
time_tolerance_in_seconds <- 60
```


### be consistent

```{r, eval = F}
# always name your test datasets with the same name, e.g. test, deleteme or df
test <- data.frame("A" = rnorm(10, 0, 1), "B" = rnorm(10, 0, 1))
deleteme <- data.frame("A" = rnorm(10, 0, 1), "B" = rnorm(10, 0, 1))
df <- data.frame("A" = rnorm(10, 0, 1), "B" = rnorm(10, 0, 1))
```


### relative over absolute file paths

Relative paths work on all computers, absolute paths only work on your computer.
```{r, eval = F}
absolute_path <- "C:/user/yourname/Documents/Project1/Rscripts/data/info_data.csv"
relative_path <- "/data/info_data.csv"

# TRICK (Additional)
# Abbreviations in paths:
# "."  : refers to the folder where you are at the moment = the working directory
# the below paths point to the same directory
path_to_data <- "./example_data_synthesis_helpdesk/"
path_to_data <- "example_data_synthesis_helpdesk/"
# ".." : refers to the parent directory (1 level up) from where you are now (=from the working directory)
#   given your path structure looks like this : 
#   C:/user/yourname/Documents/PhD/Project1/Analysis/Rscripts/data/info_data.csv
#   and your working directory is "Rscripts"
#   both of the following path points to the folder "Analysis"
path_to_data <- "../"
path_to_data <- "C:/user/yourname/Documents/PhD/Project1/Analysis/"
```


### Test your code

Testing prevents the majority of errors.
```{r, eval = F}
# create example data.frame
example <- data.frame(species = c("Erithacus rubecula", "Arion vulgaris", 
                                  "Plantago Lanceolata"),
                      abundance = c(2, NA, 10))
# action : recode NA value to 0
#    because Arion vulgaris was measured, but no individuals were found
example[which(example$species == "Arion vulgaris"), 2] <- 0
#
# test : did I do what I wanted?
sum(is.na(example$abundance)) # no NA values any more
sum(is.na(example$species))
min(example$abundance)        # min is 0 now, as expected
max(example$abundance)
View(example)
barplot(height = example$abundance, names.arg = example$species)
example$abundance
```


## Discussion

- Which of the practices will you apply (first)?
- How many of the practices did you already know, how many were new?
- Which practices, you think, will help you the most in the future?
- Which practices seem less useful to you?
