---
title: "Vegan package exploration"
author: "Caroline"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.duplicate.label = "allow")
```

```{css, echo=FALSE}
.scroll-200 {
  max-height: 200px;
  overflow-y: auto;
  background-color: inherit;
}
```

## (1) Package description

The vegan package is a very useful package  for diversity analysis and communities description. It can be found in Github at https://github.com/vegandevs/vegan/ and additional documentations on the functions can be found here : https://cran.r-project.org/web/packages/vegan/vegan.pdf 

```{r load package, message=F,warning=F, echo=T}
library("vegan") #target package
library("ape")
library("dplyr") 
library("ggplot2")
library("tidyverse")# 4 extra useful packages
```

In order to cite the package and the correct version, you can use the following command

```{r cite package, class.output="scroll-200"}
citation("vegan")
```

## (2) What can I do with the vegan package ?

#### (a) Communities descriptions

It is important to realize that the vegan package is mainly used for descriptive analyses. It is a first step in describing your communities before taking them through an efficient modelling approach. However, it cannot be used as a replacement of hypotheses ! Results obtained through ordination/non-metric multidimensional scaling (NMDS) simply describe in a lower/multi dimension(s), patterns that arise from your data.

Let's take an example and load the data that will be useful for this introduction

```{r load data, class.output="scroll-200"}
data("varespec") #this data is in the vegan package
head(varespec)
```

This data is composed of different diversity samples. For each sample (rownames), every species abundance was recorded (colnames). This data presentation is typical for communities data.

#### (b) Ordination

The first thing you can do with the vegan package is the ordination of your communities. Follow this link for an extensive introduction to ordination with vegan : https://ourcodingclub.github.io/tutorials/ordination/


```{r plot data, class.output="scroll-200"}
#direct quick output for students in a hurry...
#this is an NMDS for a specific distance calculation : Bray distance.
NMDS3 <- metaMDS(varespec, k = 2, trymax = 100, trace = F, autotransform = FALSE, distance="bray")
ef <- envfit(NMDS3, varespec, permu = 999)
ef

# The two last columns are of interest: the squared correlation coefficient and the associated p-value
# Plot the vectors of the significant correlations and interpret the plot
plot(NMDS3, type = "t", display = "sites")
plot(ef, p.max = 0.05)
```

#### (c) Cluster analysis

The second interesting thing you can do with the vegan package is a cluster analysis. The cluster analysis comes in 3 step : first, get a distance matrix, secondly, fit a cluster diagram, and thirdly, decide the optimal numbers of groups. Follow this ling for a useful tutorial on cluster analysis with vegan : https://www.introranger.org/post/cluster-analysis/

```{r load second set of data, class.output="scroll-200", warning=FALSE}
#yet another quick output if you are in a hurry...
data("varechem") #this data is in the vegan package
head(varechem)

#first step
chem_d <- select(varechem, N:Mo) #selecting interesting data
chem_m <- round(vegdist(chem_d, method="euclidean"), 1) #fit distance matrice
as.matrix(chem_m)[c(1:5), c(1:5)]#get matrix

#second step
chem_clust <- hclust(chem_m, method="ward.D2")  #cluster data
plot(chem_clust, las = 1, 
     main="Cluster diagram of soil chemistry", 
     xlab="Sample", 
     ylab="Euclidean distance")
rect.hclust(chem_clust, 2, border="red") 
rect.hclust(chem_clust, 4, border="blue")
rect.hclust(chem_clust, 5, border="darkgreen") #plot clustered data with different clustering groups

#third step
fit <- cascadeKM(chem_d, 1, 10, iter = 5000)
fit$results %>% 
        as.data.frame() %>%
        rownames_to_column("metric") %>%
          pivot_longer(names_to = "groups", 
                       values_to = "value", 
                       - metric) %>%
        mutate(groups = str_extract(groups, "\\d+"), 
               groups = as.numeric(groups)) %>%
  filter(metric != "SSE") %>%
  ggplot(aes(x=groups, y = value)) + theme_bw(16) +
  geom_line(lwd=1.5, col="blue") +
  geom_point(pch=21, col="lightgrey", 
             bg="blue", stroke = 1.5, size=5) +
  scale_x_continuous(breaks = c(2:10), labels = c(2:10)) +
  theme(panel.grid.minor.x = element_blank()) #estimate best group number for clustering
grps <- as_tibble(fit$partition) #put them in a group variable
plot(chem_clust, las = 1,
     label = grps$`5 groups`, 
  main="Cluster diagram of soil chemistry", 
  xlab="Sample", ylab="Euclidean distance")
  rect.hclust(chem_clust, 5, border="red") #final plot
```