---
title: "S2 - Data wrangling in R"
author: "N. Schenk"
date: "2022-11-09"
output: html_document
---

# Data wrangling in R

Aim of this document : give a start point to dive into data wrangling packages in R and start a discussion about it.

3 main ways to do data wrangling : 
- R base (without using a package)
- tidyverse (very user-friendly)
- data.table (for very large datasets)

Unfortunately the syntax of the 3 ways differs, and learning one way does not automatically mean learning all 3 ways. The good news is that you don't have to know all 3, but select the one that appeals most to you.

Below under `Examples` you find a few examples of classcal data wrangling problems solved with all 3 ways, for you to get an overview of all three.

## R base
Some people prefer to use base R, because it's the nearet to the programming language, and thus the quickest way, and no need to install a package before doing something. Or they prefer to learn base R before diving into the different packages in order to understand better what the functions do.

Two suggestions of tutorials to learn how to wrangel `data.frames`, the most used R object containing research data:
- [An Introduction to R by Douglas, Roos, Mancini, Couto and Lusseau](https://intro2r.com/wrangling-data-frames.html) Chapter 3.4
- In my opinion the classic of all classics : [The R book by M. Crawley](http://www.ievbras.ru/ecostat/Kiril/R/Biblio_N/R_Eng/Crawley2013.pdf)

Other possibilities are : ["R for Fledglings" by Donovan, Brown and Katz](https://www.uvm.edu/~tdonovan/RforFledglings/) starting at Chapter 6.4. and [R Guide for NSC Statistics by Deanna Li](https://bookdown.org/dli/rguide/working-with-dataset.html) Chapter 7.




## tidyverse
```{r}
# install.packages("tidyverse")
library(tidyverse)
# The tidyverse package loads a lot of packages.
#    mostly used for data.wrangling are : dplyr and tidyr
```

There are a lot of very good tutorials about this online. I recommend the book ["R for Data Science" by Hadley Wickham](https://r4ds.had.co.nz/wrangle-intro.html), where he introduces the why and how of data wrangling in Chapter 9-16.

There is a very useful [site from tydiverse](https://www.tidyverse.org/learn/) with recommendations for 2 online courses and the book mentioned above.


## data.table
```{r}
# install.packages("data.table")
library(data.table)
```
Data.table is another widely used data wrangling package, which is specialised on large datasets - it is programmed in a way that makes handling very large datasets quite fast.

A very good starting point is this [vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html).


# useful functions in R base
```{r}
# basic functions helpufl for dataset wrangling
vec <- c(1, 2, 3, 4, 5) # create example vector
vec > 3                 # show for each element of the vector if the statement is true or fals
sum(vec > 3)            # R interprets T as 1 and F as 0. We can therefore use sum() to count the number of times T occurs.
                        #    in other words, count the number of times the statement is true for our vector
which(vec > 3)          # show the indices of the vector for which the statemetn is true
                        #    in other words, the 4th and 5th element of the vector are larger than 3


# load example dataset
data(iris)


# look at the dataset
iris        # print dataset to console
View(iris)  # show dataset in a pretty way in RStudio
head(iris)  # print first 6 rows to console
tail(iris)  # print last 6 rows to console
names(iris) # print column names
nrow(iris)  # print number or rows
ncol(iris)  # print number of columns


# dataset summaries
str(iris)   # print a summary of column names, classes and values to the console
summary(iris)        # prints a summary of each column of the dataset, with mean, 1-3 quantile, min and max
unique(iris$Species) # print unique values of the column "Species" to the console, if the column e.g. contains categories
duplicated(iris$Species) # prints T or F for each value, if it is a duplicate = if it has occurred before
table(iris$Species)  # print all categories and counts of how many times a given value appears
which(iris$Sepal.Length > 6) # print the indices of all rows for which the statement is true
length(which(iris$Sepal.Length > 6)) # print the number of rows for which the statement is true


# access certain elements of a data.frame
# select columns
iris$Sepal.Length      # select a given column by name
iris[, "Sepal.Length"] # select a given column by name
iris[, 1]              # select a given column by index (the first (1) column). Note : is error prone, if you change the order
                       #    of columns in the dataset, the index will change
# select rows
iris[1, ]              # select a given row by index
iris[which(iris$Sepal.Length == 6.9), ] # select a given row by value
iris[which(iris$Sepal.Length > 6), ] # filter rows by statement using which
iris[iris$Sepal.Length > 6, ]        # filter rows by statement without using which


# perform calculations on dataset
stats::aggregate(iris, Sepal.Length ~ Species, mean) # calculate the mean Sepal.Length of each Species
apply(iris[, -which(names(iris) %in% c("Species"))], 2, mean) # calculate the mean of each column except the Species column
apply(iris[, -which(names(iris) %in% c("Species"))], 1, sum)  # calculate the sum of each row from the iris dataset without the Species column
```

Other useful functions to be potentially added : 
- from `reshape2` package : `melt()` and `dcast()` to get from long to wide dataset format 
- from `R base` : 
    - `rbind()` and `cbind()` to combine two datasets (without using a common column)
    - `merge()` to combine two datasets by a common column


# Examples
A few examples showing how to perform a given computation with all 3 proposed ways: with R base, with tidyverse and with data.table.

## Filtering a single column by name
```{r}
# with R base
iris[, "Sepal.Length"]   # output is a vector
iris$Sepal.Length        # output is a vector
iris[, c("Sepal.Length")]

# with tidyverse
tbl_iris <- tibble::as_tibble(iris)   # convert to specialised tidyverse format
dplyr::select(tbl_iris, Sepal.Length) # output is a data.frame (or tibble, the format used in tidyverse)
# note : actually the specialised data format is not absolutely needed, the commands also work on data.frames :
dplyr::select(iris, Sepal.Length) 

# with data.table
dt_iris <- data.table::data.table(iris) # need to convert example dataset to data.table format
dt_iris[, Sepal.Length]    # output is a vector
dt_iris[, .(Sepal.Length)] # output is a data.table
# note : the specialised data format is absolutely needed, the commands do not work on data.frames :
# iris[, Sepal.Length] # raises an error
```
Conclusion : all 3 ways work very well and use only short commands.


## Removing a column by name
```{r}
# filter out (remove) a column by name
iris[, -which(names(iris) %in% c("Species"))]
iris[, names(iris) != "Species"]

# with tidyverse
dplyr::select(tbl_iris, !Sepal.Length)

# with data.table
dt_iris[, -"Sepal.Length"]
```
Conclusion : The R base command is already longer than the commands from the specialised packages.


## Filtering rows
Show all rows of a data.frame that contain the category "setosa" in the column "Species".
```{r}
# with R base
iris[which(iris$Species == "setosa"), ]

# with tidyverse
dplyr::filter(iris, Species == "setosa")

# with data.table
dt_iris[Species == "setosa", ]
```


# Discussion
- Where and how do the packages differ? List aspects that you noted so far.
- Which one is most appealing to you and why?
- Try to group the differing aspects : Which ones make the package suitable for Ecology and which are rather a disadvantage?


# Rmarkdown exercise 
- Apply changes to this script in a way that the knitted html document does not contain all the very long ouputs of the iris dataset.
¨