---
title: "Advanced Data Wrangling - Solutions"
author: "N. Schenk"
date: "2023-03-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This Script contains the solutions to the tutorial about advanced data wrangling.




### Exercises "apply"

**Exercise 1** : Calculate the median number of mites in each plot in the dataset mite.
```{r}
# the mite dataset contains counds of orobatid mites of 70 soil cores (in rows) of 70 mite species.
# solution with rowMeans()
rowMeans(mite)
# solution with apply()
apply(mite, 1, mean)
```

**Exercise 2** : The varespec dataset contains plant cover data of 24 plots of 44 variables. Calculate the mean and the standard deviation of the plant cover for each plant species. Additional : store the results in a new data.frame. Additional 2 : plot the results.
```{r}
apply(varespec, 2, mean)
apply(varespec, 2, sd)

# adding output to a data.frame
specieswise_varespec <- data.frame(names = names(varespec),
           specieswise_mean = apply(varespec, 2, mean),
           specieswise_sd = apply(varespec, 2, sd))
# note the rownames are a bit irritating, they can be removed
rownames(specieswise_varespec) <- NULL

# plotting
ggplot(specieswise_varespec, aes(x = names, y = specieswise_means)) +
  geom_point() +
  geom_errorbar(aes(ymin = specieswise_means - specieswise_sd, 
                    ymax = specieswise_means + specieswise_sd)) +
  coord_flip()
#TODO : this plot still misses a nice title and nicer axis labels, and maybe some extras like colors? Would you feel to add this and send to us, so the tutorial gets nicer?
```


**Exercise 3** : The varespec dataset contains plant cover data of 24 plots of 44 variables. Calculate the estimated cover of each plot, using the apply function.

Additional : Why is the cover not always 100%?
```{r}
apply(varespec, 1, sum)
```


**Exercise 4** : During data analysis of the varespec data, it turned out that the cover data of some of the plant species need to be log transformed. Please log-transform the following columns : Cladunci Cladcocc Cladcorn Cladgrac Cladfimb Cladcris Cladchlo Cladbotr. 

*Hint* : During execution, you will run into a problem. Please try to solve as you would do in a real analysis.
```{r}
varespec_columns_to_log_transform <- c("Cladunc", "Cladcocc", "Cladcorn", "Cladgrac", "Cladfimb", "Cladcris", "Cladchlo", "Cladbotr")

apply(varespec[, which(names(varespec) %in% varespec_columns_to_log_transform)],
      2, log)
# please note that the log of 0 is not defined. Therefore, we need to take log(x + 1)

apply(varespec[, which(names(varespec) %in% varespec_columns_to_log_transform)],
      2, function(x) log(x + 1))
```


